Notes

### 2022-09-20 notes/notes.tex
optimisation problem

\documentclass{article}
\usepackage{amsmath, amssymb, dsfont, physics, hyperref, cleveref}
\usepackage[margin=1.5cm]{geometry}
\usepackage{biblatex}

@@

@@

\subsection{Model with time-dependent correction to the emissions}
If we weigh each day and each emission source by some factor, we are applying a transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,365} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, 365} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as 
$$
W (S^{-1} x \otimes \vb{1}_{365})
\quad\text{where}\quad 
W = \begin{bmatrix}
    w_{1,1} & 0 & \cdots & 0 \\
    0 & w_{1,2} & \cdots & 0 \\
    \vdots & \vdots & \ddots & 0 \\
    0 & 0 & \cdots & w_{m, 365}
\end{bmatrix}
.$$
Using this weighted vector of emissions instead of $\tilde{x}_{365}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
.\end{equation}
In this expression, the unknown is $W$ and should be fit to the data. 
There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
In the following, we provide a theoretical description of a system that can be solved for $W$.

We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices. Suppose we have a parametrisation $\Gamma: \mathbb{R}^{k} \rightarrow \mathcal{W}, u \mapsto \Gamma(u)$. 
In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
$$
\hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
.$$
If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, dsfont, physics, hyperref, cleveref}
\usepackage[margin=1.5cm]{geometry}
\usepackage{biblatex}

@@

\newtheorem{proposition}{Proposition}

@@

\subsection{Model with time-dependent correction to the emissions}
If we weigh each day and each emission source by some factor, we are applying a transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,365} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, 365} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as 
$$
W (S^{-1} x \otimes \vb{1}_{365})
\quad\text{where}\quad 
W = \begin{bmatrix}
    w_{1,1} & 0 & \cdots & 0 \\
    0 & w_{1,2} & \cdots & 0 \\
    \vdots & \vdots & \ddots & 0 \\
    0 & 0 & \cdots & w_{m, 365}
\end{bmatrix}
.$$
Using this weighted vector of emissions instead of $\tilde{x}_{365}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
.\end{equation}
In this expression, the unknown is $W$ and should be fit to the data. 
% There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
% In the following, we provide a theoretical description of a system that can be solved for $W$.

% We can put constraints on $W$ of the form $W \in \mathcal{W}$, where $\mathcal{W}$ is some set of constrained diagonal matrices. Suppose we have a parametrisation $\Gamma: \mathbb{R}^{k} \rightarrow \mathcal{W}, u \mapsto \Gamma(u)$. 
% In this case, \cref{eq: time-evolving model concise} can be reformulated as a linear operation applied to $\Gamma(u)$ as follows:
% $$
% \hat{y} = ((S^{-1} x \otimes \vb{1}_{365})^T \otimes M_{shift}) \mathrm{vec}\, \Gamma(u)
% .$$
% If $\Gamma$ is a linear map, the above is a linear expression in $u$. Then, the minimisation of $\norm{y - \hat{y}}^2$ can be solved using numerical linear algebra.

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise natural logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\frac{1}{2}\norm{F(W)}^2 + \frac{1}{2}\lambda \norm{\log W}^2
.$$
\begin{proposition}
    Let $z$ be the vector obtained by applying the function $x \mapsto (\log x) / x$ to all predicted concentrations $\hat{y}_i$ corresponding to detection events. Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left((P_{det} M_{shift})^T z \tilde{x}^T\right) + \lambda W^{-1} \log W
    $$
    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
\end{proposition}
\begin{proof}[Proof (still to be checked for calculation errors)]

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) := 
P_{det} M_{shift} W(t) \tilde{x}_{365}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}^T) U}_A t
$$
where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}^T
.
\end{align*}
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T (P_{det} M_{shift})^T z \tilde{x}^T
.$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.

The foregoing gives the derivative of the first term in $f(W)$. The gradient of the second term can be calculated in exactly the same manner: 
$$
\mathrm{grad}_W\, \left(\frac{1}{2}\norm{\log W}^2\right) = W^{-1} \log W
.$$
\end{proof}
